Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1)

### 2014-04-25 WSSSPE1_report/WSSSPE1_summary_report.tex
read and just one minor change
editing of the end of 8

Another obstacle to
sustainability identified by Elster include patenting of software.
Most countries place some limits on software patents. The European 
Union outright forbids them, while US patent law excludes ``abstract ideas,'' 
which has been used to refuse some software patents.
%
Further obstacles to
sustainability include a lack of open access, and even more broadly, a lack of open source
codes even in open access journals.  Finally, a lack of awareness on
the part of scientific software developers of commodity libraries for
common tasks reduces their ability to reuse code.

@@

Finally, it appears that the career options of developers from
two kinds of software projects differ significantly. In order to
further investigate the much discussed issue of career paths for
developers of these software, we conducted a small survey of the cases
discussed above with respect to the current status of its core developers. We
looked at the workshop submissions and the original cited papers for the
respective software. Where available, we looked at the web pages and source
code of the earliest versions to find the original developers. We included
projects older than 5 years and excluded relatively new projects such as
\toolname{SciPy} and \toolname{numFOCUS} from this survey. We find that in a
majority of scientific software, more than one original developer are still
around, in some cases after more than 10 years (e.g., \toolname{R/qtl}).
Whereas, in the case of utility software projects, it appears that at least one
original developer is still around, often in a part-time or ``benefactor''
role. In other cases, it appears that the original developers have went on to
assume a larger role (eg. manager) within the same project.

In conclusion, we find that there are distinctions in the
characteristics and the kinds of challenges faced between utility and
scientific software projects.  In terms of career paths of original
developers, it appears that the career choices for utility software
developers are broader compared to the developers of scientific
software. However, it appears that in the case of scientific software,
core competencies of original developers tend to be domain science who
happen to be the coders of their software. As projects as a whole, the
small-scale, specific software have a lower risk and cost in terms of
sustaining as compared to utility software. We also find that the
utility software tend to adhere more loyally to the many best
practices discussed during the workshop. Often the sustainability of
scientific software projects is achieved by the fact that the core
developer or team heavily utilizes the software for their science,
e.g., \toolname{R/qtl, PETSc}.  Furthermore, the development of
scientific software requires more scientific skills compared with
those of utility software. This diminishes the need of specialist
programmers and practices in many cases where the bulk of development
is carried out by a domain scientist. With these inferences, we
believe that separate guidelines and principles for sustainability are to
be defined and practiced for each category of scientific software and
they be given separate treatment.

>>>>

Another obstacle to
sustainability identified by Elster include patenting of software.
Most countries place some limits on software patents. The European 
Union outright forbids them, while US patent law excludes ``abstract ideas'', 
which has been used to refuse some software patents.
%
Further obstacles to
sustainability include a lack of open access, and even more broadly, a lack of open source
codes even in open access journals.  Finally, a lack of awareness on
the part of scientific software developers of commodity libraries for
common tasks reduces their ability to reuse code.

@@

%Finally, it appears that the career options of developers from
%two kinds of software projects differ significantly. 
%Examining the workshop submissions for 
%projects older than 5 years (excluding relatively new projects such as
%\toolname{SciPy} and \toolname{numFOCUS}) and the original cited papers for the
%respective software, and, where available, the web pages and source
%code of the earliest versions to find the original developers, it appears
%that in a majority of scientific software, more than of the one original developer are still
%associated with the project, in some cases after more than 10 years later (e.g., \toolname{R/qtl}).
%For the utility software projects, it appears that at least one
%original developer is still
%associated with the project, often in a part-time or ``benefactor''
%role, or in a larger role (e.g., manager).

In conclusion, we find that there are distinctions in the
characteristics and challenges faced between utility and
scientific software projects.  
%In terms of career paths of original
%developers, it appears that the career choices for utility software
%developers are broader compared to the developers of scientific
%software. However, it appears that in the case of scientific software,
%core competencies of original developers tend to be domain science who
%happen to be the coders of their software. 
%As projects as a whole, the
%The
%small-scale, specific software have a lower risk and cost in terms of
%sustaining as compared to utility software. 
We %also 
find that the
utility software packages are more likely to use the best
practices discussed during the workshop. Often, sustainability of
scientific software projects is achieved by the fact that the core
developer or team heavily utilizes the software for their own science,
e.g., \toolname{R/qtl, PETSc}.  Furthermore, the development of
scientific software requires more scientific background compared with
utility software, thus in many cases, the bulk of development
is carried out by a domain scientist. For these reasons, we
believe that separate guidelines and sustainability principles could
be defined for these two software categories.
